MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-15 main/MSstatsPTM_Main.tex

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates (mass spectrometry runs) per condition.  For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and the corresponding unmodified features of the protein in a special case of 2 conditions and 2 biological replicates per condition. The number of quantified features varies across replicate LC-MS/MS runs of the same sample, and across conditions. There may be feature outliers which could be caused by random variation or measurement error. Additionally, some spectral features can be missing. The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.

The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models. For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$. Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}

However, these population parameters are inherently convoluted with changes in the unmodified protein. To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$. We estimate these quantities in the same way as for the PTM, using their corresponding features and observations, resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$. These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) = 0 \\
H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) \neq 0
\end{aligned}
\label{eq:null_hyp}
\end{equation}

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

\subsection*{Existing statistical methods for experiments targeting PTMs}

\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}\cite{THOMAS2020}. While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

When modeling the simple experiment seen in Figure \ref{fig:data-structure}, we are interested in answering the Null Hypothesis in Equation \ref{eq:null_hyp}. To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively. In other words it allows us to answer the Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
Goals of PTM characterization, input to statistical analyses, and notation. (a) Schematic data representation, in a simplified case of two conditions and two biological replicates. Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities of the PTM in Condition 1 are denoted by $y_1^\ast$. Features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

>>>>

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.  For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition \todo{OV: replace 'run' with 'bio rep' in the figure}. The number of modified and unmodified features varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. We are interested in testing the null hypothesis
\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}
To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}(a), and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.

Unfortunately, the population parameter is inherently convoluted with the overall changes in protein abundance. To account for this, we can change the null hypothesis to

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) = 0 \\
H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) \neq 0
\end{aligned}
\label{eq:null_hyp}
\end{equation}
where $\mu_1$ and $\mu_2$ reflect the overall protein abundances in condition $1$ and condition $2$. These quantities can be estimated using protein features without the modification site.

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection*{Existing statistical methods for detecting differentially abundant PTMs}

\noindent {\bf Two-sample $t$-test} 
\todo{OV: could you help me with this part a little bit? I tried to organize the text, but could not quite finish. We should have the following order: (1) input, (2) model for the input, (3) model-based inference, (4) implementation; (5) good things and limitations. I think it may help to separate two versions: based on the features, and based on the ratios. Otherwise it is not very clear}
Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. $t$-test takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features \todo{OV: let's separate the intensities and the ratios}, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}. as input the adjusted PTM abundance of each condition and biological replicate, and estimates $\Delta$ in Equation \ref{eq:null_hyp}.
\todo{OV: what is the model?}
To estimate the parameters of interest in this trivial model, the approach takes as input

@@

\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations. Here good things are: complex designs, including multiple variance components; EB (for TMT); good summarization; implementation. Limitation: MSstats is not designed for protein-level adjustment of changes in PTM}

$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively. In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation. (a) Schematic data representation, in a simplified case of two conditions and two biological replicates. Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide. The feature \todo{log-}intensities of the PTM in Condition 1 are denoted by $y_1^\ast$ \todo{In the figure $y^*$ has 3 subscripts. Could you edit here as well?}. \todo{Log-intensities of } features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$ \todo{In the figure $y$ has 3 subscripts. Could you edit here as well?}. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). \todo{OV: I am wondering whether we need Fig 2(b). The null hypothesis is specified in the text. I am not sure it is referenced anywhere. Maybe remove it for simplicity?} (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}
